---
title: "matrix-computations"
author: "Ata COLAK"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

# Computations on Matrices

# Introduction

> This task shows a few examples of matrix computations using R. The task shows examples of graph plots, selecting custom rows and columns in a matrix, creating custom functions, exporting certain dates of the year to analyse, and creating custom tables.

> The dataset used is EUR currency rates until 2023-02-02, and can be accessed via the link below:

> https://raw.githubusercontent.com/gagolews/teaching-data/",
"master/marek/eurxxx-20230202-numeric.csv

## First Part - EUR-PLN Plot

> Code is explained via the comments.

```{r 1}
X <- as.matrix(read.csv(
paste0("https://raw.githubusercontent.com/gagolews/teaching-data/",
"master/marek/eurxxx-20230202-numeric.csv"), comment="#"))
head(X, 3)


# Plot of EUR-PLN for the year 2022,
# Values of "xlim" are columns where "Date" corresponds to year 2022.
# "ylim" has been increased to show the plot close-up.
plot(X[,1],X[,14] , xlim = c(18993, 19357), ylim = c(4.5, 4.95), type = "l", lwd = 1.5, ylab = "EUR-to-PLN", xlab = "Days in Unix Date")
```

## Second Part - Min, Median, Mean, Max Table

> Code is explained via the comments.

```{r 2}
# Custom function 
min_median_mean_max <- function(x) 
{
    c(min=min(x), median=median(x), mean=mean(x), max=max(x))
}


# Transpose to show the matrix the way it is in the task sheet,
# Only includes rows with "Date" less than 19357 and more than or equal to 18993,
# Using custom function min_median_mean_max,
# Columns 2 to 42 to include every currency, but not the date column.

t(apply(t(X[X[,"Date"]<19357 & X[,"Date"]>=18993,2:42]), 1, min_median_mean_max))
```

## Third Task - AUD-PLN Plot 

> Code is explained via the comments.

```{r 3}

# EUR-PLN divided by EUR-AUD gives AUD-PLN rates.

aud2pln <-  X[,14] / X[,27]

# Plot of equation:

plot(X[,1], aud2pln, xlab = "Days", ylab = "AUD-to-PLN", type = "l", lwd = 1.5)

```

## Fourth Task - Average AUD-PLN Rates for 2020, 2021, 2022

> Code is explained via the comments.

```{r 4}
# Matrix seperated in dates 2020, 2021 and 2022.

date_2020 <- X[X[,"Date"]<=18627 & X[,"Date"]>=18262,]
date_2021 <- X[X[,"Date"]<= 18992 & X[,"Date"] >=18628,]
date_2022 <- X[X[,"Date"]<=19357 & X[,"Date"]>=18993,]

# AUD-PLN rates in each year.

aud2pln2020 <- date_2020[,14] / date_2020[,27]
aud2pln2021 <- date_2021[,14] / date_2021[,27]
aud2pln2022 <- date_2022[,14] / date_2022[,27]


# Custom table with values above.
exchange_table <- matrix(c(2020, mean(aud2pln2020), 2021, mean(aud2pln2021), 2022, mean(aud2pln2022)), nrow = 3, byrow = TRUE)
colnames(exchange_table) <- c("Year", "AverageRate")

exchange_table

```

# Conclusion

> In conclusion, this task demonstrated the use of matrices in R for analysing financial data, and various methods to access certain information in matrices.

> While doing this task, I have learned how to select certain rows and columns from matrices and analyse the data I have received by putting it through certain conditions.

> I have observed that during 2020-2021-2022, the AUD-PLN rates have increased in value every year.

> What was interesting to me was the peak in EUR-PLN values around Unix date 19050. The value has increased over 5 and immediately went down again.