---
title: "data-wrangling"
author: "Ata COLAK"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---


# Data Frame Wrangling

# Introduction

> This report shows a few examples of data frame wrangling using R.

> The report shows uses of subsetting and aggregating requested information, from a requested date within a dataset, and showing it using lists and bar plots.

> The dataset used for this report is "City of Melbourne's Trees, with species and dimensions (Urban Forest)". The dataset can be accessed using the link below:

> https://data.melbourne.vic.gov.au/explore/dataset/trees-with-species-and-dimensions-urban-forest/table/ 

## First Part - Loading Dataset

> In the first example below, the csv file is read by my computer, and prints the head of the dataset to show that it has been read successfully.

```{r 1}
trees <- read.csv("trees.csv") 
head(trees, 3)
```

## Second Part - Subsetting and Aggregating the Trees Dataset to Create a Custom List

> In the second part of the code, I have created a combination for the trees of genera asked in the report sheet, to make my job easier, and to make the code easier to read.

> I first create a trees subset which only has the genus I will be using.

> Secondly, I create a new column which acts as a boolean value, if the tree was planted before 2003, it returns FALSE, if the tree is planted after 2003, it returns TRUE. I called it After2003, and I have used a conditional element selection operator, ifelse, as seen in the code.

> Afterwards, I have aggregated the subsetted list by the genera required and the After2003 check. The aggregated data also has a FUN column for the length.


```{r 2}
genera <- c("Eucalyptus", "Platanus", "Acacia", "Ulmus", "Pinus")

trees_subset <- subset(trees, Genus %in% genera)

# Create a new column for the planting period
trees_subset$After2003 <- ifelse(trees_subset$Year.Planted <= 2003,
                              "FALSE",
                              "TRUE")


# Aggregate the data by Genus and Period
trees_agg <- aggregate(trees_subset$Year.Planted,
                       by = list(Genus = trees_subset$Genus,
                                 After2003 = trees_subset$After2003),
                       FUN = length)
names(trees_agg)[3] <- "Count"

# Print the result
trees_agg
```

## Third Part - Subsetting and Aggregating the Trees Dataset to Create a Custom List and Bar Plot

> In the third part of the code, I have created another trees_subset variable trees_subset2, which also has a condition to subset trees planted after 2003.

> I have aggregated the subsetted list by genus, and FUN calculates the mean of the diameter breast height, and puts it into the second column.

> At first, I had problems creating the aggregated list, then I found the solution of giving them names by myself, that fixed the problems within the code.

> After the aggregation and naming process, I print the list, and I print the requested bar plot afterwards, with correct labels and data.

```{r 3}
trees_subset2 <- subset(trees, Year.Planted >= 2003 & Genus %in% genera)

trees_agg2 <- aggregate(trees_subset2$Diameter.Breast.Height,
                        by=list(trees_subset2$Genus), FUN=mean, na.rm = TRUE)

names(trees_agg2) <- c("Genus", "Diameter")


print(trees_agg2)
barplot(trees_agg2$Diameter, names.arg = trees_agg2$Genus, xlab = "Genus", ylab = "Average Diameter")
```

## Fourth Part - Benefits and Concerns of Sharing Such Data With the Public, and What an AI Engineer Can Do With Them

> In my opinion, sharing such detailed data about trees which extends to planting information as old as 1899 is highly beneficial for data analysts, in a way that it allows them to do whatever they want with the data. The examples are limitless, but a few would be:

> Analysing data in such way that, by going through the geolocations of the trees, a tree map of Melbourne can be created, which would show how Melbourne's nature was shaped before the trees were cut down,

> Analysing pre-dated breast height to compare how the shape changed through the course of over 100 years,

> Analysing which trees were more prevelant throughout the years...

> The options are limitless, therefore I do not see any negative concerns with sharing this dataset with the public.

# Conclusion

> In conclusion, this report demonstrated subsetting and aggregating in R for data wrangling, and I believe with my explanations, I proved there are much more different ways of handling this dataset for different purposes.

> While doing this report, I have learned how subsetting and aggregation works, and how to analyse datasets to a wider aspect.

> What was interesting to me about this report, and the thought provoking question was the amount of possibilities a creative person can see by looking at a big dataset.